---
title: "Getting and cleaning data"
output:
  html_document:
    df_print: paged
---

### Introduction
As this is quite a long exercise, I have presented my findings as an R markdown script with explanations of each step taken.

#### Loads the required dplyr and tidyr packages.
```{r}
library(dplyr)
library(tidyr)
```

---

#### Get Features  
Reads the features data file in, separating data onto new lines.  
There is no header present and all columns are converted to character data.  
The features data contains the final column names.  
Print out top 5 column headers.  
```{r}
features <- read.delim("data/features.txt", sep = "\n", colClasses = "character", header = FALSE)
features[1:5,]
```

Separates the feature data columns into index and value based on a white space separator.  
Splits data eg 1 <--> tBodyAcc-mean()-X  
```{r}
featuresDf <- separate(features, 1, c("c", "v"), sep = "(\\s\\s|\\s)")
featuresDf[1:5,]
```

Produces a logical vector isolating the features we are interested in which contain \*mean or \*std  
This is an important vector which we will use later to isolate the columns we are interested in.  
```{r}
interestedFeatures <- grepl(".*mean.*|.*std.*", featuresDf[,2])
interestedFeatures[1:5]
```

---

#### Prepare Training Data  

First we need to read in the training data/train  

Reads in the y training data set and sets the column name to activity
```{r}
y_train <- read.delim("data/train/y_train.txt",  sep = "\n", colClasses = "character", header = FALSE)
colnames(y_train) <- "activity"
y_train[1:5,]
```

Gets the training subjects data and sets the column name to subject.  
```{r}
training_subjects <- read.delim("data/train/subject_train.txt",  sep = "\n", colClasses = "character", header = FALSE)
colnames(training_subjects) <- "subject"
training_subjects[1:5,]
```
Read in the x_training data and split by a whitespace separator.  
```{r}
x_train <- read.delim("data/train/x_train.txt", sep = "\n", colClasses = "character", header = FALSE)
xtrain <- separate(x_train, 1, as.character(1:562), sep = "(\\s\\s|\\s)")
xtrain[1:5,]
```

Remove first empty column and set all column names using the features data frame.  
```{r}
xtrain <- xtrain[,2:562]
colnames(xtrain) <- featuresDf[,2]
xtrain[1:5,]
```

Select only the features that we are interested in and join y and x data sets to form a training data set.
```{r}
xtrain <- xtrain[,interestedFeatures]
training <- cbind(training_subjects, y_train, xtrain)
training[1:5,]
```

Remove old variables from workspace.
```{r}
rm(features, training_subjects, x_train, xtrain, y_train)
```

---

#### Prepare Test Data  


Read activity.
```{R}
y_test <- read.delim("data/test/y_test.txt",  sep = "\n", colClasses = "character", header = FALSE)
colnames(y_test) <- "activity"
y_test[1:5,]
```


Read test subjects
```{r}
test_subjects <- read.delim("data/test/subject_test.txt",  sep = "\n", colClasses = "character", header = FALSE)
colnames(test_subjects) <- "subject"
test_subjects[1:5,]
```


Read x_test
```{r}
x_test <- read.delim("data/test/x_test.txt", sep = "\n", colClasses = "character", header = FALSE)
xtest <- separate(x_test, 1, as.character(1:562), sep = "(\\s\\s|\\s)")
x_test[1,1]
```


Remove empty first column
```{r}
xtest <- xtest[,2:562]
xtest[1,]
```

Apply column names from the features data frame.  
Extract only the columns we are interested in (mean and standard deviation).
```{r}
colnames(xtest) <- featuresDf[,2] 
xtest <- xtest[,interestedFeatures]
xtest[1:5,]
```

Join x_test and y_test creating training
```{r}
testing <- cbind(test_subjects, y_test, xtest)
testing[1:5,]
```


Remove old variables from workspace
```{r}
rm(test_subjects, x_test, xtest, y_test, featuresDf, interestedFeatures)
```

---

#### Combine Training and Test Data

Create combined data set from training and testing
```{r}
combinded <- rbind(training, testing)
rm(training, testing)
combinded[1:5,]
```

Create activity labels data set
```{r}
activityLabels <- read.delim("data/activity_labels.txt", sep = "\n", colClasses = "character", header = FALSE)
activityLabels[1:5,]
```

```{r}
activityDf <- separate(activityLabels, 1, c("c","v"), sep = "(\\s\\s|\\s)")
rm(activityLabels)
activityDf[1:5,]
```


Create Dplyr data tables
```{r}
activities <- tbl_df(activityDf)
combo <- tbl_df(combinded)
rm(combinded, activityDf)
```

Join activity labels to the main data table.  
Rename the appended column to activityLabel
```{r}
joined <- left_join(combo, select(activities, c, v), by = c("activity" = "c"))
colnames(joined)[82] <- "activityLabel"
rm(combo, activities)
joined[1:5,]
```

Replace activity with its related label
```{r}
joined <- joined %>% mutate(activity = activityLabel)
joined[1:5,]
```

Remove temporary activityLabel column added earlier.
```{r}
joined <- select(joined, -c("activityLabel"))
joined[1:5,]
```

Convert all readings columns to the numeric type.
```{r}
joined <- joined %>% mutate_at(3:81,as.numeric)
joined[1:5,]
```

Group data by subject and activity.
```{r}
grouped <- joined %>% group_by(subject, activity)
```

Summarise the data columns by the mean function.
```{r}
result <- grouped %>% summarise_each(mean)
```

Print out final data table.
```{r}
rm(joined, grouped)
result
```
